---
title: "US Opioid Prescriber Analysis"
output: html_document
date: '2022-06-18'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Preparation

```{r, include=FALSE}
#load required libraries
library(tidyr)
library(data.table)
library(dplyr)
library(tidyverse)
library(arules)
```

The original data sets used for this project can be downloaded from the following websites: 
Medicare Part D Prescribers - by Provider and Drug:   https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers/medicare-part-d-prescribers-by-provider-and-drug/data

Medicare Part D Prescribers - by Geography and Drug: 
https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers/medicare-part-d-prescribers-by-geography-and-drug/data

We also referenced code in Kaggle script below for preparing the final data set for further analysis: https://raw.githubusercontent.com/apryor6/apryor6.github.io/master/Identifying-Opioid-Prescribers/create-dataset.R.

A copy of the final data set to a csv file and used in the following analysis.

### Exploratory and Descriptive Analysis

```{r}
# read final prescriber data set created above
prescriber <- read.csv('prescriber_2019.csv')
prescriber_final <- prescriber
str(prescriber_final)
```

```{r}
#check if the data set is complete
nrow(prescriber_final[!complete.cases(prescriber_final),]) 
#0 means there are no incomplete rows
```

```{r}
#convert prescriber NPI to factor
prescriber_final$PRSCRBR_NPI <- as.factor(list(prescriber_final$PRSCRBR_NPI))

#convert character variables to factors
prescriber_final <- prescriber_final %>% 
  mutate_if(is.character, funs(as.factor))

#convert output variable to factor
prescriber_final$OPIOD_PRESC <- as.factor(prescriber_final$OPIOD_PRESC)
```

First, we reviewed how many prescriber types are included in the data set.
```{r}
str(prescriber_final$PRSCRBR_TYPE)
```
As shown above, the data set includes 183 different prescriber types.

We then checked how many states are included.
```{r}
str(prescriber_final$PRSCRBR_STATE_ABRVTN)
```

The data set includes information for 61 "states", which are 50 United States, District of Columbia, U.S. territories, Armed Forces areas, Unknown and Foreign Country. For our analysis, we will remove the Unknown and Foreign Country records and use information of the remaining 59 "states".

```{r}
# remove Unknown and Foreign Country records that are belong to states XX and ZZ
prescriber_final_new<-prescriber_final[!(prescriber_final$PRSCRBR_STATE_ABRVTN %in% c("XX","ZZ")),]

prescriber_final_new$PRSCRBR_STATE_ABRVTN <- as.character(prescriber_final_new$PRSCRBR_STATE_ABRVTN)
prescriber_final_new$PRSCRBR_STATE_ABRVTN <- as.factor(prescriber_final_new$PRSCRBR_STATE_ABRVTN)

str(prescriber_final_new$PRSCRBR_STATE_ABRVTN)
```

```{r}
#review the number of records in the new data set
nrow(prescriber_final_new)
```

We also created a subset of data that only includes opiate drugs to conduct exploratory and descriptive analyses.

```{r}
#select only basic information and data for opiate drugs
prescriber_final_slim <- prescriber_final_new[,c("PRSCRBR_STATE_ABRVTN", "PRSCRBR_TYPE", "ACETAMINOPHEN.WITH.CODEINE",  "MORPHINE.SULFATE", "TRAMADOL.HCL","HYDROMORPHONE.HCL", "METHADONE.HCL", "FENTANYL", "OXYCODONE.HCL.ACETAMINOPHEN", "HYDROCODONE.ACETAMINOPHEN", "OXYCODONE.HCL","TOPIRAMATE","OPIOD_PRESC")]

# create another subset that only includes opiate prescriber
opioid_prescriber <- prescriber_final_slim[prescriber_final_slim$OPIOD_PRESC  == TRUE,]

# create another subset that only includes non-opiate prescriber
non_opioid_prescriber <- prescriber_final_slim[prescriber_final_slim$OPIOD_PRESC  == FALSE,]
```

```{r}
# calculate the proportion of the non-opioid precriber
nrow(non_opioid_prescriber)/nrow(prescriber_final_slim)
```
Around 59% of the prescribers never prescribed opiate drugs before. 

```{r}
#add new column that summarizes the total opioids prescriptions
opioid_prescriber$TOTAL_PRESCRIPTIONS <- rowSums(opioid_prescriber[,c(3:12)])
```


```{r}
#check the statistics of the Total Prescriptions variable
summary(opioid_prescriber$TOTAL_PRESCRIPTIONS)
```

Per results above, the distribution of total number of opioid prescriptions written by opioid prescribers is extremely right skewed with a median of 49.0 and mean of 152.6.

We wanted to check the top 10 opioid prescribers by total number of prescriptions.

```{r}
opioid_prescriber[,c('PRSCRBR_STATE_ABRVTN','PRSCRBR_TYPE','TOTAL_PRESCRIPTIONS')] %>%
  arrange(-TOTAL_PRESCRIPTIONS) %>%
  head(10)
```

As shown above, an individual from LA issued more that 30,000 total number of opioid prescriptions in 2019, which was significantly more than all other prescribers in the entire US.

```{r}
#Group the prescriber data by Prescriber Type
groupedByPrescriber <- opioid_prescriber %>% 
  group_by(PRSCRBR_TYPE) %>% 
  summarise(Total_Prescriber_Count=n(), Total_Opioids_Count = sum(TOTAL_PRESCRIPTIONS)) %>%
  arrange(-Total_Opioids_Count)

#check the top prescriber types with highest total prescription amounts
groupedByPrescriber %>%
  arrange(-Total_Opioids_Count) %>%
  head()
```

The top three prescriber types by total number of opioid prescriptions were Family Practice, Internal Medicine and Nurse Practitioner.

We then checked what were the top three prescriber types by average number of opioid prescriptions. The average was calculated from total number of prescriptions divided by total number of prescribers.
```{r}
#create a column for average number of opioids prescription per prescriber
groupedByPrescriber$Avg_Opioids_Count <- round(groupedByPrescriber$Total_Opioids_Count / groupedByPrescriber$Total_Prescriber_Count)

groupedByPrescriber %>%
  arrange(-Avg_Opioids_Count) %>%
  head()
```

This time, we identified another unusual case. Durable medical equipment & medical supplies specialty had only one opioid prescriber in 2019, but the average number of opioid prescriptions this person wrote ranked the fourth among all specialties. 	 

We created a visualization for top 10 prescriber_types which had the highest number of opioid prescribers.

```{r}
bp_groupedByPrescriber <- groupedByPrescriber[1:10,] %>%
  mutate(PRSCRBR_TYPE = fct_reorder(PRSCRBR_TYPE, -Total_Prescriber_Count)) %>%
  ggplot(aes(x=PRSCRBR_TYPE, y=Total_Prescriber_Count)) +
  geom_bar(stat="identity", fill="lightblue")+
  theme_minimal()+
  ggtitle("Total Number of Opioid Prescribers By Specialty") + 
  labs(x = "Prescriber Specialty", y = "Prescriber Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

bp_groupedByPrescriber
```

We then reviewed the prescriber and prescription data by State.
```{r}
#summarize the total number of prescriptions for each opioid by State
groupedByState <- aggregate(cbind(ACETAMINOPHEN.WITH.CODEINE,  MORPHINE.SULFATE, TRAMADOL.HCL,HYDROMORPHONE.HCL, METHADONE.HCL, FENTANYL, OXYCODONE.HCL.ACETAMINOPHEN, HYDROCODONE.ACETAMINOPHEN, OXYCODONE.HCL,TOPIRAMATE,TOTAL_PRESCRIPTIONS)~PRSCRBR_STATE_ABRVTN, data = opioid_prescriber, sum)

groupedByState %>%
  arrange(-TOTAL_PRESCRIPTIONS) %>%
  head()
```

The top three States by total number of opioid prescriptions were CA, FL and TX.


### Association Rule Mining

**Data Preprocessing**

We first performed the Association Rule Mining analysis. The goal of the Association Rule Mining is to explore relations between Prescriber Type and Drug information, more specifically which opioids are most likely to be prescribed by certain types of medical professionals.  Therefore, we conducted the research using only opioid prescribers and opiate drugs data.

First, we reviewed the opioid prescriber data again.
```{r}
summary(opioid_prescriber)
```

As shown above, all drug variables are numeric and are extremely right skewed (means are far greater than the medians). Therefore, we would like to convert them to factors by discretizing those variables into customized bins.  
```{r}
#create a function to discretize drug counts into customized bins
split_drugcount <- function(name,df){
df[,name] <-  cut(df[,name], breaks = c(-1,0,10,100,1000,2000,4000,Inf),labels=c("0", "0-10","10-100","100-1000","1000-2000","2000-4000", "Above 4000"))
return(df)
}

#create another function to discretize drug counts into different customized bins
split_drugcount2 <- function(name,df){
df[,name] <-  cut(df[,name], breaks = c(-1,10,500,1000,2000,4000,Inf),labels=c( "0-10","10-500","500-1000","1000-2000","2000-4000", "Above 4000"))
return(df)
}

#create another function to discretize drug counts into different customized bins
split_drugcount3 <- function(name,df){
df[,name] <-  cut(df[,name], breaks = c(-1,10,100,500,1000,Inf),labels=c( "0-10","10-100","100-500","500-1000","Above 1000"))
return(df)
}

```

We tried using different customized bins, and found out the second one above performed the best results. Therefore, the following analysis will be conducted by applying that discretization function. We also tried to discretize those drug variables into equal-length bins, but found out more meaningful rules could be generated if using customized bins.

```{r}
#save a copy of the data in a new variable
opioid_prescriber_arm <- opioid_prescriber

#create the data set that only includes prescriber state, prescriber type and opiate drugs
opioid_prescriber_arm <- opioid_prescriber_arm[,1:12]
str(opioid_prescriber_arm)
```

```{r}
#perform discretization by applying the function split_drugcount2 to the data
final_opiate_drugs_list <- list('ACETAMINOPHEN.WITH.CODEINE', 'MORPHINE.SULFATE', 'TRAMADOL.HCL', 'HYDROMORPHONE.HCL', 'METHADONE.HCL', 'FENTANYL', 'OXYCODONE.HCL.ACETAMINOPHEN', 'HYDROCODONE.ACETAMINOPHEN', 'OXYCODONE.HCL', 'TOPIRAMATE')

for (name in final_opiate_drugs_list) {
  opioid_prescriber_arm <- split_drugcount2(name,opioid_prescriber_arm)
}

#review the new data frame
summary(opioid_prescriber_arm)
```
The final data frame only includes information for 401,790 opioid prescribers.

**Data mining**
Next, we performed association rule mining on the data set using the apriori function in the arules package. We started with generating rules that have a minimum support of 0.10, and a minimum confidence of 0.8.
```{r}
a <- Sys.time()

#generate rules using arules package and apriori function
rules <- apriori(opioid_prescriber_arm, parameter = list(supp = 0.10, conf = 0.80, minlen = 2,maxlen=2))

difftime(Sys.time(),a)
```

Rules were generated in a second.

```{r}
# Show the rules that have a lift value greater or equal to a value
goodrules <- rules[quality(rules)$lift > 1]

# show the rules that have prescriber type on the left hand side
rules_subset <- subset(goodrules, (lhs %in% paste0("PRSCRBR_TYPE=", unique(opioid_prescriber_arm$PRSCRBR_TYPE))))
inspect(rules_subset)
```

As shown above, only three rules that had prescriber type on the left hand side had a lift value greater than 1. Then, we adjusted the support and confidence thresholds to include more rules.

```{r}
a <- Sys.time()

rules2 <- apriori(opioid_prescriber_arm, parameter = list(supp = 0.03, conf = 0.70, minlen = 2,maxlen=2))

difftime(Sys.time(),a)
```

```{r}
# Show the rules that have a lift value greater or equal to 1
goodrules2 <- rules2[quality(rules2)$lift > 1]

# show the rules that have prescriber type as left hand side
rules_subset2 <- subset(goodrules2, (lhs %in% paste0("PRSCRBR_TYPE=", unique(opioid_prescriber_arm$PRSCRBR_TYPE))))
inspect(rules_subset2)
```

After reviewing the 41 rules generated above, we found 5 most interesting rules and they were put in the final report.

We also did additional research to check if we can find relations between State and Drug information.

```{r}
a <- Sys.time()

rules3 <- apriori(opioid_prescriber_arm, parameter = list(supp = 0.01, conf = 0.70, minlen = 2,maxlen=2))

difftime(Sys.time(),a)
```

```{r}
# Show the rules that have a lift value greater or equal to a value
goodrules3 <- rules3[quality(rules3)$lift > 1]

# show the rules that have prescriber type as left hand side
rules_subset3 <- subset(goodrules3, (lhs %in% paste0("PRSCRBR_STATE_ABRVTN=", unique(opioid_prescriber_arm$PRSCRBR_STATE_ABRVTN))))
inspect(rules_subset3[1:30])
```

Unfortunately, no interesting rules were generated based on results above. Association rule mining may not be a good technique to analyze relations between State and Drug Name/Prescription Amount.


### Conclusion
The rules identified suggest that if the prescriber is practicing family practice, he or she is around 81% likely to prescribe Tramadol 10-500 times a year, and around 74% likely to prescribe Hydrocodone Acetaminophen 10-500 times a year. If the prescriber is practicing internal medicine, he or she is around 78% likely to prescribe Tramadol 10-500 times a year.  If the prescriber is practicing orthopedic surgery or emergency medicine, he or she is around 73% likely to prescribe Hydrocodone Acetaminophen 10-500 times a year.

The results from the Association Rule Mining may be used by the government authority to monitor which specialties are most likely to prescribe opioids at high dosages. We also reviewed rules that have the State name on the left-hand side but did not find any interesting results. Association Rule Mining may not be a good technique to analyze relations between State and drug name/prescription amount.

